Validation of marker gene expression

### 2025-02-27 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
add back heatmap and skip 24

```{r, results='asis'}
# check for if the project is all cell lines or not 
is_cell_line <- all(consensus_dt$sample_type == "cell line")

@@

>>>>

```{r, results='asis'}
# check for if the project is a cell line project or not 
# explicitly skip project 24 since the current processed data doesn't have the updated combination of PDX and cell line
is_cell_line <- all(consensus_dt$sample_type == "cell line") | params$project_id == "SCPCP000024"

@@

The heatmap shows the mean expression of each marker gene across all cells assigned to a given cell type group. 
In contrast to the dot plot, we will show all marker genes for a given cell type group, not just the uniquely expressed genes.

```{r, eval=!is_cell_line}
# add column to uniquely label gene symbols since some genes are present across multiple cell types 
# add this here since we will use the order in this data frame to set the order of the heatmap columns
heatmap_markers_df <- filtered_markers_df |> 
  dplyr::mutate(
    unique_symbol = glue::glue("{gene_symbol}-{validation_group_annotation}")
  ) 
```

```{r, eval=!is_cell_line}
# first make a mtx to use for the heatmap with rows as cell types and genes as columns 
heatmap_mtx <- group_stats_df |> 
  dplyr::ungroup() |> 
  # add unique label for gene symbol 
  dplyr::mutate(
    unique_symbol = glue::glue("{gene_symbol}-{validation_group_annotation}")
  ) |> 
  dplyr::select(unique_symbol, broad_celltype_group, mean_exp) |> 
  dplyr::filter(unique_symbol %in% heatmap_markers_df$unique_symbol) |> 
  tidyr::pivot_wider(
    names_from = unique_symbol,
    values_from = mean_exp
  ) |> 
  tibble::column_to_rownames("broad_celltype_group") |>
  as.matrix()
# replace any NA with 0
# accounts for marker genes that aren't found in all cells 
heatmap_mtx[is.na(heatmap_mtx)] <- 0
# make sure cell types are present in the right order
heatmap_mtx <- heatmap_mtx[c(celltype_order, "unknown"), heatmap_markers_df$unique_symbol]

# create annotation for heatmap
annotation <- ComplexHeatmap::columnAnnotation(
  group = heatmap_markers_df$validation_group_annotation,
  col = list(
    group = celltype_colors
  ),
  annotation_legend_param = list(
      title = "Validation group"
    )
)
```

```{r, eval=!is_cell_line, fig.height=7, fig.width=10}
marker_gene_heatmap(heatmap_mtx, annotation)
```
